### 2017-02-08 manuscript.rmd
word smithing

There is now considerable interest in comparing functional genomics data across species. One goal of this work is to provide an integrated understanding of genome and phenotype evolution. Most studies of this type have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species. The statistical problems that arise from not considering these relationships can lead pairwise approaches to the wrong conclusions, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare gene expression across species with pairwise methods. We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique features of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor. This violates the fundamental assumption of observation independence in standard statistical methods. Phylogenetic comparative methods address this dependence. They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics. Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Figure 1a). This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1b). The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue by re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

@@

The second comparative functional genomics study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. Using pairwise comparisons, they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Figure 1a). They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla. We previously described concerns with the interpretations of this result [@Hejnol:2016hw]. Here we address the analyses themselves by examining the structure of the pairwise comparisons.

![Pairwise and phylogenetic comparative approaches, illustrated on an example gene tree with multiple genes per species. The internal nodes of the tree are speciation and gene duplication events. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison dependent on others. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.pdf)

@@

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that have not been investigated in detail. It can be applied to any trait of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

The KMRR study considered several publicly available datasets of gene expression across tissues and species. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. This includes information about which tissue expression is specific to. For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.

@@

For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes. One of these trees is shown in Supplementary Figure 7 to demonstrate the analysis.

@@

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree. The other reason for the different conclusions is that pairwise comparisons between the tips and phylogenetic comparative methods describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure 4). There are several difficulties with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

These simulations have several implications. The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture even when the evolution of expression is `r dup_adjust` fold higher following duplication than speciation. The pairwise results are the same under both hypotheses. They also serve as a validation of phylogenetic methods as applied to this problem. Our analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture. In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hypotheses about evolutionary process.

@@

Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species. We found that all five of the lowest values in the mid phase distribution (Figure 6a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 6b). This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1a). This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other. Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].

>>>>

There is now considerable interest in comparing functional genomics data across species. One goal of this work is to provide an integrated understanding of genome and phenotype evolution. Most studies of this type have relied on multiple pairwise comparisons of functional genomic data between species, an approach that does not incorporate information about the evolutionary relationships among species. The statistical problems that arise from not considering these relationships can lead pairwise approaches to the wrong conclusions, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare gene expression across species with pairwise methods. We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique features of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods has been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor. This violates the fundamental assumption of observation independence in standard statistical methods. Phylogenetic comparative methods address this dependence. They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics. Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1a). This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1b). The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny. One reason that comparative functional genomic sudies have not embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue by re-evaluating the pairwise comparisons in two recent studies that compared gene expression across species.

@@

The second comparative functional genomics study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. Using pairwise comparisons, they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw]. Furthermore, they suggested that this provides biological justification for the concept of phyla. We previously described concerns with the interpretations of this result [@Hejnol:2016hw]. Here we address the analyses themselves by examining the structure of the pairwise comparisons.

![Pairwise and phylogenetic comparative approaches, illustrated on an example gene tree with multiple genes per species. The internal nodes of the tree are speciation and gene duplication events. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes across species. This leads to many more comparisons than unique observations, making each comparison dependent on others. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomic studies.](Figure_overview.pdf)

@@

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is also used to relate findings from well-studied genes to related genes that have not been investigated in detail. It can be applied to any trait of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

The KMRR study considered several publicly available datasets of gene expression across tissues and species. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). Tau is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. This includes information about which tissue expression is specific to. For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.

@@

For each internal node in each gene tree, we calculated the phylogenetic independent contrast (PIC) of Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication).

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned)` gene trees passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes. One of these trees is shown in Supplementary Figure 7 to demonstrate the analysis.

@@

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree. The other reason for the different conclusions is that pairwise comparisons and phylogenetic comparative methods describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. Any difference could be due to the structure of the gene phylogenies alone. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs. We confirmed that we can replicate this result (Figure 4). There are several difficulties with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

These simulations have several implications. The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture even when the evolution of expression is `r dup_adjust` fold higher following duplication than speciation. The pairwise results are strikingly similar under both hypotheses. These simulations also serve as a validation of phylogenetic methods as applied to this problem. Our phylogenetic analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture. In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hypotheses about evolutionary process.

@@

Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species. We found that all five of the lowest values in the mid phase distribution (Figure 6a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 6b). This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1a). This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other. Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].
